The Bolocam Galactic Plane Survey: Survey Description and Data Reduction

### 2009-04-16 methods/ms.tex
Minor corrections to paper
cygnus comparison figure updated

% -----------------------------------------------------------------------------
% BGPS specific definitions
\def\bgps{BGPS}
\newcommand{\bgpsarea}{150}
\newcommand{\bcamfwhm}{31.2\arcsec}
\newcommand{\ncores}{7628}
\newcommand{\bgpsdepthlow}{20}
\newcommand{\bgpsdepthhigh}{50}
% -----------------------------------------------------------------------------
% Missions
\def\spitzer{{\em Spitzer}}
% -----------------------------------------------------------------------------

@@

The data were reduced with a custom pipeline written specifically for
A key element of the reduction is the iterative estimation of the
atmospheric fluctuations and the astrophysical signal.  The
atmospheric model is developed from a set of principal components of
the bolometer signal time series under the assumption that the bulk of
the correlated signal is atmospheric.  The iterative estimation
process then helps to restore correlated flux due to astrophysical
emission (Section \ref{sec:Mapping}).

The Bolocam observations are sensitive to structure between 31\arcsec\
and about 400\arcsec\ (the array field of view).  An indicative
analysis of the spatial filter function of the iterative mapping
scheme is given in \citet{enoch06}.  The data were sampled onto a
uniform 7\farcs2 grid.  Flux density calibrations were performed using
observations of Mars, Uranus, and Neptune, and they are accurate to
$\sim$20\%.  This is described in \ref{sec:FluxCalibration}.

%A calibration scheme capable of
%5\% systematic and $<$5\% statistical errors is being implemented as outlined in
%\citet{Sayers2009}.
%To account for offsets between the observations, each observation was mapped
%individually and aligned to a `master' field that has been tied to
%pointing calibrators. 
Noise properties across the survey are not completely uniform, but the
measured $1\sigma$ level is $\approx43\pm7$ mJy/beam within the
$b=\pm.5$ portion of the plane.
%After the pipeline mapping, the fields were combined using the COMBINE
%command in IRAF with WCS offsets.

The absolute pointing accuracy of the survey has been measured to be
$\sim$6\arcsec\ RMS.  A single $3^\circ \times 1^\circ$ observation of
each region was registered to the pointing calibrators as a ``master
field'', and additional observations of that field were aligned to the
master (Section \ref{sec:Astrometry}).

Bright and large sources leave an impact on their surroundings when
cleaned with a principal component subtraction.  \object{Sgr B2},
\object{G 34.3+0.15}, and other sources with fluxes $\geq 10$ Jy are
surrounded by substantial negative bowls.  Some flux is restored in
these bowls using the iterative process 
%described in \citet{Aguirre2009}
, but around the brightest sources the impact of the source on the
background estimation is too great to be fully recovered (Section
\ref{sec:Photometry}).

Instrument glitches and cosmic ray hits have been removed by a
combination of outlier rejection in each spatial pixel and hand
flagging of the timestream data.  Streaking in the images results from
incorrect determinations of the zero level across scans when a bright
source is encountered.  Atmospheric noise in observations taken in
poor weather may not be completely removed by the reduction
procedures (Section \ref{sec:Noise}).
%Particular bolometers may be noisy for a single observation and they may not
%be adequately down-weighted.

@@

To relate observations of the primary calibrators to observations of
the BGPS fields, we make use of the following relation.  The
calibration ${\cal S}$, referred to the detectors, is given by
\begin{equation}
\label{eq:Calibration}
{\cal S} \, \left[\mathrm{\frac{V}{Jy}}\right] = 
S(\tau) \eta A \exp{(-\tau)} \Delta\nu
\end{equation}
where $S$ is the bolometer responsivity ([V/W]), $\eta$ is the system
optical efficiency, $A$ is the effective telescope collecting area,
$\Delta \nu$ is the bandwidth, and $\tau$ is the line-of-sight,
in-band atmosphere opacity.  Under the assumption that the only power
variation on the detectors is due to the power from the atmosphere,
which may be parameterized by $\tau$, ${\cal S}$ is a single-valued
function of $\tau$.  We have used the mean bolometer resistance as a
proxy for $\tau$, since the bolometer resistance is a single-valued
function of loading.  {\it This quantity is monitored continuously for
all observations.}  Note that this calibration curve folds in both the
effects of changing atmospheric transmission, as well as changes in
the detector response with optical loading.

@@

A fit to the observed values for the primary calibrators was performed
for the each epoch separately.
%CHECK
The agreement between epochs was good, so a single combined
calibration was used for all data here.
%We have found that the calibration can differ
%between successive cooldowns of the instrument by $\sim10\%$ \Check\
%but is generally stable over the period of order a month which
%constitutes each of the observing epochs.  
The calibration curve is shown in Figure \ref{fig:CalibrationCurves}.  The
resulting error on the calibration curve fit is less than 
%CHECK
$5\%$ (statistical) over the observed range of $\tau$.  A tally of all
contributions to the flux density error is given in Table
\ref{tab:ErrorBudget}.

@@

The instrument beam size (PSF) is measured using the planets Uranus
and Neptune which are nearly point sources for Bolocam.
%The semi-diameter of
%Mars over this period ranged from 2.3\arcsec\ to 2.9\arcsec, with
%Neptune $\sim1.1\arcsec$.  Relative to the \bcamfwhm~FWHM Bolocam
%beam, these may be regarded as point sources; the finite size of Mars
%results in a $< 1.5\%$ correction of the observed flux.
The maps of the planets, including all detectors, are fit to a
Gaussian profile.  The beam area is that averaged over all detectors,
and is ?$1.71\times10^{-6}$\TBD\ ($35.6$\farcs) steradians.  This
allows conversion of the maps, calibrated in Jy beam$^{-1}$, into
surface brightness (MJy sr$^{-1}$).  The uncertainty in the beam size
is ?$.7\%$\TBD\ (number quoted is the uncertainty in the mean beam
size), leading to an uncertainty in the surface brightness calibration
of \TBD.

@@

The figures of the Cygnus-X region presented by M07
%Motte et al. (2009)
, obtained at a central wavelength of 1.2 mm using the 30 meter IRAM
telescope, appear nearly identical to the 1.1 mm Bolocam images.  The
BGPS mapping strategy, larger beam, and field of view provides better
recovery of the large scale structure of the dust continuum emission,
but inferior angular resolution.  Nevertheless, there is excellent
agreement between the structure revealed by the two data sets.

The Bolocam and M07 data sets have flux calibration consistent within
about 20\%.  However, direct comparison with the fluxes listed in M07
is difficult due to the factor of three difference in the size of the
beams used.  For example, the the brightest source in the Cyg-X region
is DR 21.  While M07 quote a peak flux density per beam of
4.2 Jy/beam for the brightest feature in the DR 21 complex (their peak
N46) and 17.9 Jy/beam in a 0.19 x 0.14 pc box, BGPS measures a peak
flux of 17.5 Jy/beam (the beam being roughly 35\arcsec\ in diameter).
Fluxes are best compared using sources compact as observed with the
beam of the IRAM 30 meter telescope.

@@

% 7
\Figure{cygnus_flux_comparison}{A comparison of the fluxes extracted
in the same aperture from the \citet{motte07} MAMBO map and the Bolocam map of
this paper. THIS FIGURE IS A PLACEHOLDER}{fig:CygnusComparison}{1.0}

@@

\Figure{image_types}{Examples of four of the image types being released.  Top
left: MAP.  Top right: Residual.  Bottom left: Deconvolved.  Bottom right:
Weight}{fig:sampleimages}{1.0}

>>>>

% -----------------------------------------------------------------------------
% BGPS specific definitions
\def\bgps{BGPS}
\newcommand{\bgpsarea}{150}
\newcommand{\bcamfwhm}{35.6\arcsec}
\newcommand{\ncores}{7628}
\newcommand{\bgpsdepthlow}{20}
\newcommand{\bgpsdepthhigh}{50}
% -----------------------------------------------------------------------------
% Missions
\def\spitzer{{\em Spitzer}}
% -----------------------------------------------------------------------------

@@

The data were reduced with a custom pipeline written specifically for the BGPS.
A key element of the reduction is the iterative estimation of the
atmospheric fluctuations and the astrophysical signal.  The
atmospheric model is developed from a set of principal components of
the bolometer signal time series under the assumption that the bulk of
the correlated signal is atmospheric.  The iterative estimation
process then helps to restore correlated flux due to astrophysical
emission (Section \ref{sec:Mapping}).

The Bolocam observations are sensitive to structure between 35\arcsec\ and
about 200\arcsec\ (half the array field of view, see figure
\ref{fig:PCA_filter}).  An indicative analysis of the spatial filter function
of the iterative mapping scheme is given in \citet{enoch06}.  The data were
sampled onto a uniform 7\farcs2 grid.  Flux density calibrations were performed
using observations of Mars, Uranus, and Neptune, and they are accurate to
$\sim$20\%.  The flux calibration is described in section \ref{sec:FluxCalibration}.

%A calibration scheme capable of
%5\% systematic and $<$5\% statistical errors is being implemented as outlined in
%\citet{Sayers2009}.
%To account for offsets between the observations, each observation was mapped
%individually and aligned to a `master' field that has been tied to
%pointing calibrators. 
Noise properties across the survey are not completely uniform, but the
measured $1\sigma$ level is $\approx43\pm7$ mJy/beam within the
$b=\pm.5$ portion of the plane in the 1st quadrant.
%After the pipeline mapping, the fields were combined using the COMBINE
%command in IRAF with WCS offsets.

The absolute pointing accuracy of the survey has been measured to be
$\sim$6\arcsec\ RMS (figure \ref{fig:PointingModel}.  A single $3^\circ \times
1^\circ$ observation of each region was registered to the pointing calibrators
as a ``master field'', and additional observations of that field were aligned
to the master (Section \ref{sec:Astrometry}).

Bright and large sources leave an impact on their surroundings when
cleaned with a principal component subtraction.  \object{Sgr B2},
\object{G 34.3+0.15}, and other sources with fluxes $\geq 10$ Jy are
surrounded by substantial negative bowls.  Some flux is restored in
these bowls using the iterative process, 
 but around the brightest sources the impact of the source on the
background estimation is too great to be fully recovered (Section
\ref{sec:Photometry}).

Instrument glitches and cosmic ray hits have been removed by a combination of
outlier rejection in each spatial pixel and hand flagging of the timestream
data.  Streaking in the images results from incorrect determinations of the
zero level across scans when a bright source is encountered and variable noise
in each strip from variable dwell time.  Atmospheric noise in observations
taken in poor weather may not be completely removed by the reduction procedures
because the atmospheric opacity can vary over timescales comparable to the time
to scan astrophysical sources (Section \ref{sec:Noise}).

@@

To relate observations of the primary calibrators to observations of
the BGPS fields, we make use of the following relation.  The
calibration ${\cal S}$, referenced to the detectors, is given by
\begin{equation}
\label{eq:Calibration}
{\cal S} \, \left[\mathrm{\frac{V}{Jy}}\right] = 
S(\tau) \eta A \exp{(-\tau)} \Delta\nu
\end{equation}
where $S$ is the bolometer responsivity ([V/W]), $\eta$ is the system
optical efficiency, $A$ is the effective telescope collecting area,
$\Delta \nu$ is the bandwidth, and $\tau$ is the line-of-sight,
in-band atmosphere opacity.  Under the assumption that the only power
variation on the detectors is due to the power from the atmosphere,
which may be parameterized by $\tau$, ${\cal S}$ is a single-valued
function of $\tau$.  We have used the mean bolometer resistance as a
proxy for $\tau$, since the bolometer resistance is a single-valued
function of loading.  {\it This quantity is monitored continuously for
all observations.}  Note that this calibration curve folds in both the
effects of changing atmospheric transmission and the changes in
the detector response with optical loading.

@@

A fit to the observed values for the primary calibrators was performed
for the each epoch separately.
%CHECK
The agreement between epochs was good, so a single combined
calibration was used for all data.
%We have found that the calibration can differ
%between successive cooldowns of the instrument by $\sim10\%$ \Check\
%but is generally stable over the period of order a month which
%constitutes each of the observing epochs.  
The calibration curve is shown in Figure \ref{fig:CalibrationCurves}.  The
resulting error on the calibration curve fit is less than 
%CHECK
$5\%$ (statistical) over the observed range of $\tau$.  A tally of all
contributions to the flux density error is given in Table
\ref{tab:ErrorBudget}.

@@

The instrument beam size (PSF) is measured using the planets Uranus
and Neptune which are nearly point sources for Bolocam.
%The semi-diameter of
%Mars over this period ranged from 2.3\arcsec\ to 2.9\arcsec, with
%Neptune $\sim1.1\arcsec$.  Relative to the \bcamfwhm~FWHM Bolocam
%beam, these may be regarded as point sources; the finite size of Mars
%results in a $< 1.5\%$ correction of the observed flux.
The maps of the planets, including all detectors, are fit to a
Gaussian profile.  The beam area is that averaged over all detectors,
and is $1.71\times10^{-6}$ steradians ($35.6$\farcs).  This
allows conversion of the maps, calibrated in Jy beam$^{-1}$, into
surface brightness (MJy sr$^{-1}$).  The uncertainty in the mean beam size
is $.7\%$, leading to an uncertainty in the surface brightness calibration
of \TBD.

@@

The figures of the Cygnus-X region presented by \citet{motte07}, obtained at a
central wavelength of 1.2 mm using the 30 meter IRAM telescope, appear nearly
identical to the 1.1 mm Bolocam images.  The BGPS mapping strategy, larger
beam, and field of view provides better recovery of the large scale structure
of the dust continuum emission, but inferior angular resolution.  The MAMBO
filter is wider than the Bolocam filter and includes the CO 2-1 line.
Nevertheless, there is excellent agreement between the structure revealed by
the two data sets.

The Bolocam and M07 data sets have flux calibration consistent within about
20\%.  However, direct comparison with the fluxes listed in M07 is difficult
due to the factor of three difference in the size of the beams used.  For
example, the the brightest source in the Cyg-X region is DR 21.  While M07
quote a peak flux density per beam of 4.2 Jy/beam for the brightest feature in
the DR 21 complex (their peak N46) and 17.9 Jy/beam in a 0.19 x 0.14 pc box,
BGPS measures a peak flux of 17.5 Jy/beam (the beam being roughly 35\arcsec\ in
diameter).  Fluxes are best compared using sources that appear compact as
observed with the beam of the IRAM 30 meter telescope.

@@

% 7
\Figure{cygnus_flux_comparison}{A comparison of the fluxes extracted
in the same aperture from the \citet{motte07} MAMBO map and the Bolocam map of
this paper. }{fig:CygnusComparison}{1.0}

@@

\Figure{image_types}{Examples of four of the image types being released.  Axes are offsets
in arcminutes.  See section \ref{sec:FinalMaps}. Top
left: Map.  Top right: Residual.  Bottom left: Deconvolved.  Bottom right:
Weight}{fig:sampleimages}{1.0}
